Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-19 co-regular-EPQ.tex
modify

Let $\Gamma\mathsf{\text{-}Condition}(e)$ be $\mathsf{HasEmptyRemoves}(e)$, $\neg \mathsf{HasEmptyRemoves}(e) \wedge \mathsf{HasUnmatchedMaxPriority}(e)$, and $\mathsf{HasMatchedMaxPriority}(e)$, for $\Gamma = \mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}$, respectively. $\textit{Check-PQ-Conc}$ inspires us that, whenever $\Gamma\mathsf{\text{-}Condition}(e)$ holds, there should exist some $\alpha$, such that  $\Gamma\mathsf{\text{-}Conc}(e,\alpha)$ holds. The following lemma states this formally. Let $\textit{proj}(e)$ be the set of projections of $e$ into set of values. When refer to $\textit{proj}(e)$, we implicitly assume that each $\textit{rm}(\textit{empty})$ in $e$ has a ghost argument that is unique.

@@

Register automata is a kind of finite automata with registers used to deal with infinite data domain, where register can be assigned values and do simple operations. In this paper we use a restricted kind of register automata, which contains only one register, whose value is obtained once by guessing and never changed. Our register machine contains a register $\textit{reg}$. Its transition labels are $\textit{call}(\textit{put},d,g), \textit{ret}(\textit{put},d,p),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards. A guard is a predicate chosen from $\{\textit{equalReg},\textit{lesReg},\top \}$, which represents equals the value of $\textit{reg}$, less than the value of $\textit{reg}$, and always holds, respectively. Given an execution $e = \alpha_1 \cdots \alpha_k$ of priority queue, register automata $\mathcal{A}$ accepts $e$, if there exists value $p \in \in \mathbb{D}$ and transitions $q_0 \xrightarrow{\beta_1} q_1 \ldots \xrightarrow{\beta_k} q_k$ of $\mathcal{A}$ from a initial state to a final state, such that for each $i$

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $\alpha_i = \textit{call}(\textit{put},a,p)$, then $\beta_i= \textit{call}(\textit{put},a,g)$ and $g(q)$ holds.

@@

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
$\seqPQ$ is co-regular, if for each $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$, there is a finite set $\textit{Auts}_{\Gamma}$ of register automata, such that for each data-independence implementation $\mathcal{I}$, we have that:

$\textit{Auts}_{\Gamma} \cap \mathcal{I} \neq \emptyset$ if and only if $\exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\Gamma\mathsf{\text{-}Condition}(e')$ holds but $\Gamma\mathsf{\text{-}Conc}(e',\alpha)$ does not hold for some $\alpha$
\end{definition}

@@

Given a data-independent implementations $\mathcal{I}$ of extended priority queue, $\mathcal{I} \cap \textit{Auts}_{\textit{sinPri}} \neq \emptyset$, if and only if there exists $e \in \mathcal{I}_{\neq}$, $e' \in \textit{proj}(e)$, such that $e'$ is single-priority  without $\textit{rm}(\textit{empty})$, and $\textit{transToQueue}(e')$ does not linearizable to queue.
\end{restatable}

@@

The second point is to ignore the incomparable priorities. Given a data-differentiated execution $e$, we say that $e$ is a $p$-execution, if $p$ is larger than or equal to the priority of values of $e$. Given a data-differentiated execution $e$ and assume that $p$ is one of the maximal priority of $e$, let $\textit{p-Exec}(e)$ be an execution obtained from $e$ by erasing all operations of values whose priority are incomparable with $\textit{pri}$. The following lemma states that when checking co-regular of $\seqPQ$, it is safe to consider only $p$-executions.

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Given a data-differentiated execution $e$ and $\Gamma\in \{\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$. $\Gamma\mathsf{\text{-}Conc}(e,x)$ holds, if and only if $\Gamma\mathsf{\text{-}Conc}(\textit{p-Exec}(e),x)$ holds, where $p$ is the priority of $x$.
\end{restatable}
\begin {proof} (Sketch)

@@

Given $\textit{p-Exec}(e) \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \Gamma\mathsf{\text{-}Seq}(s,\alpha)$. By above lemma we know that $e$ has a linearization $l'$, such that the projection of $l'$ into $p$-incomparable values. It is not hard to see that the priorities comparable with $p$ are either smaller or incomparable with the priorities incomparable with $p$. Thus, $l' \in \seqPQ$. \qed
\end {proof}

Lemma \ref{lemma:pri execution is enough} make our co-regular proof more clear. One consequence of it is that for $e=\textit{p-Exec}(e)$, at most one $\Gamma\mathsf{\text{-}Condition}(e')$ holds.

\subsection{Co-Regular of $\textit{EPQ}_1^{>}$}
\label{subsec:co-regular of EPQ1Lar}

In this subsection, we briefly introduce the idea for proving co-regular of $\textit{PQ}_1^{>}$. The proof of this subsection can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. The notion of left-right constraint used for extended priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}.

Given a data-differentiated $\_$-execution $e$ such that $\textit{last}(e) = \textit{EPQ}_1^{>}$, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensuring that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$. This is because that it is possible that $e \not\sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$ because of interaction between actions of multiple priorities. One such example is shown in \figurename~\ref{fig:introduce gap for EPQ1Lar}. We call the time interval from $\textit{ret}(\textit{put},x)$ to $\textit{cal}(rm,x)$, or from $\textit{ret}(\textit{put},x)$ when $\textit{cal}(rm,x)$ does not exist, the interval of item $x$. In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each item by dashed line. Here we assume that $p_1 <_{\mathbb{P}} p_4$, $p_2 <_{\mathbb{P}} p_4$ and $p_3 <_{\mathbb{P}} p_4$. The reason of $e \not\sqsubseteq \textit{MS}(\textit{PQ}_1^{>})$ is that, each time point from $\textit{cal}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is in interval of some item with smaller priority.

>>>>

Let $\Gamma\mathsf{\text{-}Condition}(e)$ be $\mathsf{HasEmptyRemoves}(e)$, $\neg \mathsf{HasEmptyRemoves}(e) \wedge \mathsf{HasUnmatchedMaxPriority}(e)$, and $\mathsf{HasMatchedMaxPriority}(e)$, for $\Gamma = \mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}$, respectively. $\textit{Check-PQ-Conc}$ inspires us that, when $\Gamma\mathsf{\text{-}Condition}(e)$ holds, there should exist some $\alpha$, such that  $\Gamma\mathsf{\text{-}Conc}(e,\alpha)$ holds. The following lemma states this formally. Let $\textit{proj}(e)$ be the set of projections of $e$ into set of values. When refer to $\textit{proj}(e)$, we implicitly assume that each $\textit{rm}(\textit{empty})$ in $e$ has a ghost argument that is unique.

@@

Register automata is a kind of finite automata with registers used to deal with infinite data domain, where register can be assigned values and do simple operations. In this paper we use a restricted kind of register automata, which contains only one register, whose value is obtained once by guessing and never changed. Our register machine contains a register $r$. Its transition labels are $r=*,\textit{call}(\textit{put},d,g), \textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards. The first transition from initial state is a $r=*$ transition, and is used to assign an arbitrary value to $r$. A guard is a predicate chosen from $\{=r,\leq r,\top \}$, which represents equals the value of $\textit{reg}$, less than the value of $\textit{reg}$, and always holds, respectively. Given an execution $e = \alpha_1 \cdots \alpha_k$ of priority queue, register automata $\mathcal{A}$ accepts $e$, if there exists value $d_r\in \mathbb{D}$ and transitions $q_0 \xrightarrow{\beta_1} q_1 \ldots \xrightarrow{\beta_k} q_k$ of $\mathcal{A}$ from a initial state to a final state, such that for each $i$

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $\alpha_i = \textit{call}(\textit{put},a,p)$, then $\beta_i= \textit{call}(\textit{put},a,g)$ and $g$ holds with $p$ and $d_r$. Similarly for $\textit{ret}(\textit{put},a,p)$.

@@

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
Given $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$. $\Gamma$ is co-regular, if there is a finite set $\textit{Auts}_{\Gamma}$ of register automata, such that for each data-independence implementation $\mathcal{I}$, we have that:

$\textit{Auts}_{\Gamma} \cap \mathcal{I} \neq \emptyset$ if and only if $\exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\Gamma\mathsf{\text{-}Condition}(e')$ holds but $\Gamma\mathsf{\text{-}Conc}(e',\alpha)$ does not hold for some $\alpha$.

$\seqPQ$ is co-regular, if each $\Gamma$ is co-regular. 
\end{definition}

@@

Given a data-independent implementations $\mathcal{I}$ of extended priority queue, $\mathcal{I} \cap \textit{Auts}_{\textit{sinPri}} \neq \emptyset$, if and only if there exists $e \in \mathcal{I}_{\neq}$, $e' \in \textit{proj}(e)$, such that $e'$ is single-priority  without $\textit{rm}(\textit{empty})$, and $\textit{toQueue}(e')$ does not linearizable to queue.
\end{restatable}

@@

The second point is to ignore the incomparable priorities. Given a data-differentiated execution $e$ and one of its maximal priority $p$, $e\vert_{\preceq p}$ is an execution obtained from $e$ by erasing operations of items with priority incomparable with $p$. The following lemma states that when checking co-regular of $\seqPQ$, it is safe to ignore all incomparable priorities.

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Given a data-differentiated execution $e$ and $\Gamma\in \{\mathsf{UnmatchedMaxPriority}$, $\mathsf{MatchedMaxPriority}\}$. $\Gamma\mathsf{\text{-}Conc}(e,x)$ holds, if and only if $\Gamma\mathsf{\text{-}Conc}(e\vert_{\preceq p},x)$ holds, where $p$ is the priority of $x$.
\end{restatable}
\begin {proof} (Sketch)

@@

Given $e\vert_{\preceq p} \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \Gamma\mathsf{\text{-}Seq}(s,\alpha)$. By above lemma we know that $e$ has a linearization $l'$, such that the projection of $l'$ into $p$-incomparable values. It is not hard to see that the priorities comparable with $p$ are either smaller or incomparable with the priorities incomparable with $p$. Thus, $l' \in \seqPQ$. \qed
\end {proof}

Lemma \ref{lemma:pri execution is enough} makes our co-regular proof more clear. One consequence of it is that for $e=\textit{p-Exec}(e)$, at most one $\Gamma\mathsf{\text{-}Condition}(e')$ holds.

\subsection{Co-Regular of $\mathsf{MatchedMaxPriority}^{>}$}
\label{subsec:co-regular of EPQ1Lar}

To facilitate the proof of co-Regular of $\mathsf{MatchedMaxPriority}$, we separate $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ into the following two sub-predicates, where the first one represents that $x$ is the only value with priority $p$ in $e$ , and the second one represents that there are more than one value with priority $p$ in $e$. Here $p\not\prec'\textit{priorities}(l)$ means that $p$ is either larger, equal, or incomparable with priorities in $l$, and $p$ must equals with some priority in $l$.

{\small
\begin{align*}
\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $p\not\preceq \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4mm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4mm}\mbox{and $p\not\in \textit{priorities}(v\cdot w)$} \\
\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Seq}(e,x)=\mathsf{true} & \mbox{ iff  $e= u\cdot \textit{put}(x,p)\cdot v\cdot \textit{rm}(x)\cdot w$, $p\not\prec' \textit{priorities}(u\cdot v\cdot w)$,} \\
&\hspace{4mm}\mbox{$p\not\preceq \textit{unmatched-priorities}(u\cdot v\cdot w)$, $\textit{matched}_\prec(u\cdot v,p)$,} \\
&\hspace{4mm}\mbox{and $p\not\in \textit{priorities}(v\cdot w)$} 
\end{align*}
}

The corresponding conditions are defined as follows: $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$

{\small
\begin{align*}
\mathsf{HasEmptyRemoves}(e)=\mathsf{true} & \mbox{ iff  $e$ contains a $\textit{rm}(\textit{empty})$ operation} \\
\mathsf{HasUnmatchedMaxPriority}(e)=\mathsf{true} & \mbox{ iff $p\in \textit{unmatched-priorities}(e)$ for a maximal priority} \\
&\hspace{4mm}\mbox{$p\in priorities(e)$}
\end{align*}}

$\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ is obtained from $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Conc}(e,x)$ by using $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Seq}(e,x)$ instead of $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$. And similarly for $\mathsf{MatchedMaxPriority}^{=}\mathsf{\text{-}Conc}(e,x)$.

In this subsection, we briefly introduce the idea for proving co-regular of $\mathsf{MatchedMaxPriority}^{>}$. The proof of this subsection can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. The notion of left-right constraint used for priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}.

$\Gamma\mathsf{\text{-}Condition}(e')$ holds but $\Gamma\mathsf{\text{-}Conc}(e',\alpha)$ does not hold

Given a data-differentiated $\_$-execution $e$ such that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Condition}(e)$ holds, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensuring that $\mathsf{MatchedMaxPriority}^{>}\mathsf{\text{-}Conc}(e,x)$ for some $x$. This is because that it is possible that $e \not\sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$ because of interaction between actions of multiple priorities. One such example is shown in \figurename~\ref{fig:introduce gap for EPQ1Lar}. We call the time interval from $\textit{ret}(\textit{put},x)$ to $\textit{cal}(rm,x)$, or from $\textit{ret}(\textit{put},x)$ when $\textit{cal}(rm,x)$ does not exist, the interval of item $x$. In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each item by dashed line. Here we assume that $p_1 <_{\mathbb{P}} p_4$, $p_2 <_{\mathbb{P}} p_4$ and $p_3 <_{\mathbb{P}} p_4$. The reason of $e \not\sqsubseteq \textit{MS}(\textit{PQ}_1^{>})$ is that, each time point from $\textit{cal}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is in interval of some item with smaller priority.
